---
title: "Jaxon's Cyclistics Capstone"
author: "Jaxon Song"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Install Basic Package for Reading & Data Analysis

```{r}
library(tidyverse)
library(readr)
```

##Read Dataset

```{r read}
jan21 <- read.csv("202101-divvy-tripdata.csv")
feb21 <- read.csv("202102-divvy-tripdata.csv")
mar21 <- read.csv("202103-divvy-tripdata.csv")
apr21 <- read.csv("202104-divvy-tripdata.csv")
may21 <- read.csv("202105-divvy-tripdata.csv")
jun21 <- read.csv("202106-divvy-tripdata.csv")
jul21 <- read.csv("202107-divvy-tripdata.csv")
aug21 <- read.csv("202108-divvy-tripdata.csv")
sep21 <- read.csv("202109-divvy-tripdata.csv")
oct21 <- read.csv("202110-divvy-tripdata.csv")
nov21 <- read.csv("202111-divvy-tripdata.csv")
dec21 <- read.csv("202112-divvy-tripdata.csv")
```

##Processing & Cleaning
###Cleaning package installation and loading

```{r clean pack}
library(dplyr)
library(skimr)
library(janitor)
library(mice)
```

##RBind Dataset & Empty Pattern Analysis
```{r combine data}
year2021 <- rbind(jan21,feb21,mar21,apr21,may21,jun21,jul21, aug21, sep21, oct21, nov21, dec21)
```

##Drop NA for Distance and Station Analysis
```{r drop null}
ds_year2021 <- year2021 %>%
  drop_na(start_station_name, start_station_id, end_station_id, end_station_name, end_lat, end_lng)

```

###trim the tables, make the format clear without extra
###spaces, ds_tables are for distance and station analysis
###and tdt_tables are for time, duration, and bike type prefer
```{r cleannames}
tdt_year2021 <-clean_names(year2021)
ds_year2021 <-clean_names(ds_year2021)
```

###Replace Empty Cells with NA for tdt analysis
```{r empty drop}
tdt_year2021[tdt_year2021 == "" | tdt_year2021 == " "] <-NA
ds_year2021[ds_year2021 == "" | ds_year2021 == " "] <-NA
```

###Clean Duplicates
```{r distinct col}
tdt_year2021 <- distinct(tdt_year2021)
ds_year2021 <- distinct(ds_year2021)
```

###Use ymd_hms() to convert character date string into dates for time and 
###duration analysis
```{r  time transfer}
library(lubridate)
tdt_year2021$started_at <- ymd_hms(tdt_year2021$started_at)
tdt_year2021$ended_at <- ymd_hms(tdt_year2021$ended_at)
```

###Including Plots Packages
```{r graphing packs}
library(ggplot2)
library(ggrepel)
```

# I. discover the affect of preferred bike types
```{r Preferred Bikes, echo=FALSE}
#Extract Month
tdt_inmonth_year2021 <- tdt_year2021 %>%
  mutate(month = format(tdt_year2021$started_at, "%m"))

#tdt_inmonth_year2021 <- filter(tdt_inmonth_year2021, member_casual == "member")
ggplot(tdt_inmonth_year2021, aes(x = month, fill = member_casual)) +
  geom_bar(position = "dodge") +
  geom_text(aes(label = ..count..), stat = "count", hjust = 0.5, 
            colour = "black", size = 2.3, 
            position = position_dodge(width = 0.9)) +
  facet_wrap(~rideable_type) +
  labs(title = "2021_Bike_Preferred_member",
       y = "number_of_bikes") +
  scale_fill_manual(values=c("darkturquoise","red")) +
  coord_flip()
```

#Analysis II. Duation Analysis
```{r duration, echo=FALSE}

tdt_duration_year2021 <- tdt_inmonth_year2021 %>%
  mutate(duration = difftime(ended_at, started_at, units = "mins")) %>%
  mutate_if(is.difftime, round, digits = 3)
tdt_duration_year2021$duration <- as.numeric(tdt_duration_year2021$duration)

duration_plot2021 = tdt_duration_year2021 %>%
  group_by(month, member_casual) %>%
  summarise(number_customers = n(), avg_riding_time = round(mean(duration),  digits = 3))

options(dplyr.summarise.inform = FALSE)

duration_plot2021 %>%
  ggplot(aes(x = month, y = avg_riding_time, fill= member_casual, group = member_casual)) +
  geom_line(aes(linetype = member_casual)) +
  geom_point(shape=21, size=3) +
  geom_text_repel(aes(label = avg_riding_time))+
  labs(title = "2021_Bike_Preferred_member", 
       y = "avg_riding_time (min)") + 
  scale_fill_manual(values=c("green","red"))
```

#Analysis III. Time Analysis
```{r Time, echo=FALSE}

time_year2021 <- tdt_year2021 %>%
  mutate(time = format(tdt_year2021$started_at, "%H"))
         
time_plot2021 <- time_year2021 %>%
  group_by(time, member_casual) %>%
  summarise(user_at_the_time = n())

time_plot2021 %>%
  ggplot(aes(x = time, y = user_at_the_time, fill = member_casual)) +
  geom_bar(stat = "identity", position=position_dodge())

time_plot2021 %>%
  ggplot(aes(x = time, y = user_at_the_time, fill = member_casual, group = member_casual)) +
  geom_line(aes(linetype = member_casual)) +
  geom_label_repel(aes(label = user_at_the_time)) +
  geom_point(shape=21, size=3) +
  labs(title = "2021_Time_Analysis", x = "Time of the Day")
```

#Analysis IV. Distance Analysis
```{r Distance Analysis, echo=FALSE}
#4. Distance Analysis
library(geosphere)
ds_distance_year2021 <- ds_year2021 %>%
  mutate(distance = round(distVincentyEllipsoid
                          (ds_year2021[, c("start_lng", "start_lat")],
                            ds_year2021[, c("end_lng", "end_lat")])/1000, 3))

#Extract to month and plot
ds_distance_year2021$started_at <- ymd_hms(ds_distance_year2021$started_at)
ds_inmonth_year2021 <- ds_distance_year2021 %>%
  mutate(month = format(ds_distance_year2021$started_at, "%m"))

distance_month2021 <- ds_inmonth_year2021 %>%
  group_by(month, member_casual) %>%
  summarise(number_customers = n(), avg_distance = round(mean(distance),  digits = 2))

distance_month2021 %>%
  ggplot(aes(x = month, y = avg_distance, fill= member_casual, group = member_casual)) +
  geom_line(aes(linetype = member_casual)) +
  geom_point(shape=21, size=3) +
  geom_label_repel(aes(label = avg_distance))+
  labs(title = "2021_average_distance", 
       y = "avg_distance (km)")
```

#Analysis V. Station Analysis

###Top10 start_station for casual
```{r Most Popular Casual, echo=FALSE}

ds_startid2021 <- ds_year2021 %>%
  group_by(start_station_id, start_station_name, member_casual)  %>%
  summarise(number_customers = n()) %>%
  filter(start_station_id != 'NA', member_casual == 'casual') %>%
  arrange(-number_customers)


#Top10 end_station for casual
ds_endid2021 <- ds_year2021 %>%
  group_by(end_station_id, end_station_name, member_casual)  %>%
  summarise(number_customers = n()) %>%
  filter(end_station_id != 'NA', member_casual == 'casual') %>%
  arrange(-number_customers)

#The Overall Top10 most Popular Station for Casuals
ds_startid2021 <- ds_startid2021 %>%
  rename(station_id = start_station_id, station_name = start_station_name)
ds_endid2021 <- ds_endid2021 %>%
  rename(station_id = end_station_id, station_name = end_station_name)  
ds_station2021 = ds_startid2021 %>%
  inner_join(ds_endid2021, by  = "station_name") 

ds_casual_station <- ds_station2021 %>%
  mutate(number_of_customers = number_customers.x + number_customers.y) %>%
  rename(station_id = station_id.x, member_casual = member_casual.x) %>%
  select(station_id, station_name, member_casual, number_of_customers) %>%
  arrange(-number_of_customers)

head(ds_casual_station, 10)
```

###Top10 start_station for members
```{r Most Popular Members}
ds_member_startid2021 <- ds_year2021 %>%
  group_by(start_station_id, start_station_name, member_casual)  %>%
  summarise(number_customers = n()) %>%
  filter(start_station_id != 'NA', member_casual == 'member') %>%
  arrange(-number_customers)

#Top10 end_station for members
ds_member_endid2021 <- ds_year2021 %>%
  group_by(end_station_id, end_station_name, member_casual)  %>%
  summarise(number_customers = n()) %>%
  filter(end_station_id != 'NA', member_casual == 'member') %>%
  arrange(-number_customers)

#The Overall Top10 most Popular Station for members
ds_member_startid2021 <- ds_member_startid2021 %>%
  rename(station_id = start_station_id, station_name = start_station_name)
ds_member_endid2021 <- ds_member_endid2021 %>%
  rename(station_id = end_station_id, station_name = end_station_name)  
ds_member_station2021 <- ds_member_startid2021 %>%
  inner_join(ds_member_startid2021, by  = "station_name") 

ds_member_station <- ds_member_station2021 %>%
  mutate(number_of_customers = number_customers.x + number_customers.y) %>%
  rename(station_id = station_id.x, member_casual = member_casual.x) %>%
  select(station_id, station_name, member_casual, number_of_customers) %>%
  arrange(-number_of_customers)
 
head(ds_member_station, 10)
```
